Naturalness in U(1) extensions of the MSSM

### 2015-01-20 paper/e6ssm-tuning-paper.tex
More foxes of typos, rephrasings and soem significant edit of conclusions section.

In this paper we will examine this source of fine tuning in more
detail by considering low energy constructions where the usual fine
tuning problem from the Higgs is minimised.  We will also consider
alternate charges for the extra $U(1)$ symmetry to relax the focus on
the E$_6$SSM and demonstrate that this is quite a generic result.

@@

= \underset{i}{\max}\,\{\Delta_{p_i}\}$.\\ \\

@@

\begin{figure}[h]
\begin{center}
\resizebox{!}{6cm}{%
\includegraphics{mssm_stop_tuning.png}}
\resizebox{!}{6cm}{%
\includegraphics{mssm_higgs_tuning.png}}
\caption{Left panel: Scatter plot of fine tuning in the MSSM as a function of
the lightest stop mass, $m_{\tilde{t}_1}$, for the cut-off scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV. 
Right panel: Scatter plot of fine tuning in the MSSM as a function of
the lightest Higgs mass, $m_{h_1}$, for the cut-off scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV.}
\label{Fig:mssmstoptuning}
\end{center}
\end{figure}
To illustrate this, in the left panel of Fig.~\ref{Fig:mssmstoptuning}
we show the variation in fine tuning when we scan over the stop masses
and mixing, with $500 \textrm{ GeV} \leq m_{Q_3},m_{u_3}\leq
10\textrm{ TeV}$ and $-3810\textrm{ GeV}\leq A_t\leq -20\textrm{
  GeV}$, for $M_X=20$ TeV, $50$ TeV and $100$ TeV.  The remaining
parameters we fix such that at $M_{SUSY}$ they have the values
$\mu=-97.5$ GeV, $B=-84.8$ GeV, $M_1=92.1$ GeV, $M_2=95.9$ GeV,
$M_3=352$ GeV, $A_b=-117.9$ GeV, $A_\tau=-7.8$ GeV, $m_{L_i}=400$ GeV,
$m_{e_i}=204$ GeV, $m_{Q_{1,2}}=438$ GeV, $m_{u_{1,2}}436$ GeV and
$m_{d_i}=438$ GeV ($i=1,2,3$). Although we should stress that making
this choice will lead to a spectrum which is in conflict with the LHC
limits, doing so ensures that fine tuning due to the other parameters
is small, so that we avoid washing out the fine tuning impact of the
stops when the tuning is small\footnote{For models in which the
  spectrum is heavier, when the stop masses are small the fine tuning
  reaches a lower bound imposed by other heavier parameters.} as can
be the case when the stop masses are less than 1 TeV.  Note that the
Higgs mass is also allowed to vary in this scan, as shown in the right
panel of Fig.~\ref{Fig:mssmstoptuning}. This illustrates the tuning
problem which people have been worrying about since the discovery of
the 125 GeV Higgs boson as we see that raising the stop masses is also
pushing up the Higgs mass, meaning that heavier Higgs masses require
more fine tuning.  However for a low value of the UV scale this tuning
is not so severe unless the stops are very heavy, and a 125 geV Higgs
can be obtained without much tuning in this unrealistic case where we
have minimised other sources of tuning.  On the other hand the tuning
becomes more severe as we increase the cutoff such that for $M_X =
10^{16}$ GeV a lightest stop masses of $1-3$ TeV can result in a fine
tuning of $\approx 100 - 1000$ and the minimum tuning we see consistent
with a $125$ GeV Higgs is $\approx 200$, as shown in
Fig.~\ref{Fig:mssmstoptuning}.

@@

Since the stop mass does not have such a large impact on the fine
tuning when the cutoff scale is very low we can use this to see more
clearly the impact of the $Z^\prime$ mass on fine tuning. To do so we
select a fixed low cutoff of $M_X=20$ TeV and compare the fine tuning
for two different values of the $Z^\prime$ mass. We choose to look at
$M_{Z^\prime}=2.5$ TeV, which is just above the current limits, and
$M_{Z^\prime}=4.5$ TeV, which should be in reach in run II at the LHC
and then compare the fine tuning calculated in each case to the tuning
in the MSSM.  For this, we have performed a large six dimensional
parameter space scan in both the MSSM and E$_6$SSM, varying those
parameters most relevant for the fine tuning and the Higgs mass.
Therefore the set of parameters which we vary includes $\mu$, $B$ and
$\tan \beta$ for the MSSM, and $\lambda$, $A_\lambda$ and $\tan
\beta$, for the E$_6$SSM, which appear at tree level in the EWSB
conditions of the models. While the contribution from large stop
masses to the fine tuning is small for such a low cut-off scale, the
stop contributions to the effective potential can play a significant
role in reducing the fine tuning.  For this reason it is still
important to properly treat the tuning associated with stop
contributions to the one-loop effective potential, and so we also scan
over the soft masses $m_{Q_3}^2$, $m_{u_3}^2$ and the stop mixing
$A_t$.  The relevant parameters and ranges that were scanned over are
summarised in Table \ref{tab:scanranges}. With the exception of the
stop mixing $A_t$ and the soft trilinear $A_\lambda$, which are
logarithmically scanned, all parameters are linearly scanned.  In
addition to this we also repeat each scan for three different values
of $M_2$ to allow more variation in the chargino masses.

@@

Nonetheless the MSSM the impact of chargino limit is shown in
Fig.~\ref{Fig:chargino-plot}.  There we see that if the the full
  parameter space with $m_{\chi^\pm} < 700$ GeV was excluded the
  impact would be to make the tuning in the MSSM with a $20$ TeV
  cutoff slightly worse than that of the E$_6$SSM with the same cutoff
  and a $Z^\prime$ mass just larger than current limits.  would be
  similar to that can also be attcked more directly by looking at
  chargino mass limits as this rasies $\mu$ directly.  In the
  E$_6$SSM, while raising the chargino limit can have the same impact
  in principle, due to current limits on the $Z^\prime$ mass already
  imposing a significant degree of tuning, chargino masses do not make
  much of a noticable change.

@@

Interestingly the coefficient $d$ is very small (and negative) in the
case of the $U(1)_I$.  This allows a dramatic reduction in the fine
tuning from the $U(1)_I$ symmetry.  This is a result of the $H_u$
charge associated with $U(1)_I$ vansihing, which means that the
D-terms to the lightest Higgs which is predominantly $H_u$ at large
$\tan \beta$ are suppressed, making it difficult to raise the Higgs
mass in the same way as happens in the other models and explaining why
% DH:: I strongly suspect that they are invalid - but running scan
% to check this currently
the heavier Higgs values are less well populated [can't be obtained? -
are we sure those scattered points are valid?]  than in the other
models.  Therefore the fine tuining behaviour in this model is closer
to that of the MSSM, though the LHC can still constrain the most
natural scenarions from just placing tougher limits on the $Z^\prime$
mass.

@@

To further illustrate this point we looked at how the tuning varies
with $M_X$ for low tuning benchmarks in the MSSM and E6SSM. These
benchmarks are defined in table \ref{} and the results are shown in
Fig.~\ref{Fig:BMs-varyMX}. Since the behaviour is quite complicated we
now discuss these in detail as it provides some insight into the many
differences in the tuning between the two models.

@@

By contrast the tuning for E$_6$SSM BM2 is very different.  This point
was chosen as it had much lighter stops and a gluino mass that is just
above the experimental limit of $1.4$ TeV \cite{}.  At $20$ TeV it was
not amongst the lowest tuned points, since at that scale the tree
level tuning from $M_{Z^\prime}$ dominates.  The reduction in $M_3$
means that $\Delta_{M_3}$ is substantially lower and only becomes the
dominant tuning very close to the GUT scale, giving a tuning at
$10^{16}$ GeV of $\approx 900$, which is below that of the other two
benchmark points.  Since large mixing was also used to reduce allow
lighter stops, we see initially a large contribution from
$\Delta_{A_t}$ as we did in the MSSM benchmark.  [Last thing that needs to be explained - why do $\Delta_{M_3}$ and $\Delta_{m_{H_u}^2}$ start to decrease at large $M_X$.  I'm guessing something funny to do with their values and also  $m_{Q}^2$ and $M_3$ at the highscales.  but i need to see the numbers to know.]

@@

Prior to stringent experimental constraints on the mass of the
lightest Higgs boson and squarks in supersymmetric models, a simple
picture of a natural SUSY model emerged from theoretical reasoning,
with soft masses set to similar values at the GUT scale through local
gravity interactions with the hidden sector.  Through the use of
renormalisation group running one can then see that at the EW scale
the stops and soft Higgs masses (which appear directly in the EWSB
conditions) should appear in the the EWSB condition for $M_Z$ and
therefore it was expected that these masses should not be much bigger
than $100$ GeV.  However to disturb this elegant picture first LEP
placed constraints on the Higgs mass, requiring it to be above 114.4
GeV \cite{}, which already introduced significant tuning for
constrained models,and then the LHC measured the Higgs to be around
$125$ GeV, since heavy stops are required to raised the lightest
Higgs mass above it's tree level bound.

@@

In contrast we find that in $U(1)$ extensions of the MSSM there is an
additional way to constrain the naturalness of the models, which is
through the $Z^\prime$ mass limit.  We find when we impose a low
cutoff of $20$ TeV for setting the soft masses, the lowest tuning in
the E$_6$SSM is compatible with a $Z^\prime$ mass of $2.5$ TeV was
$\Delta \approx 100$, while if the LHC run II can place a limit of
$4.5$ TeV on $M_Z^\prime$ then the tuning would be just under $400$.
By comparison the current situation in the MSSM only requires a tuning
of around $20-30$. [Dylan: update numbers a bit -- I suck at reading
  them off the graphs:)] This should be interpreted as saying that in
the most conservative limits one can place on naturalness on these
models, the tuning in the E$_6$SSM is worse.  However if the chargino
limits blah blah blah then the situation in the two models would be
the same.

This should also be contrasted with what happens as we raise the
cutoff scale.  We showed that when the cutoff reaches $M_X =
[insert]$ the tuning of the MSSM becomes larger than that of the
E$_6$SSM.  Therefore the conclusion is not that one model is less
tuned than another, but rather that the answer depends on what
assumptions are made about the SUSY breaking.

Finally we also looked at the tuning in different $U(1)$ extensions for
fixed $\tan \beta = 10$.  We found that in all cases the fine tuning
was much worse for the larger $Z^\prime$ mass, further emphasising
the importance they play in these $U(1)$ extensions.  Interestingly
the actual level in tuning varied quite a bit between these models
with the $U(1)_I$ model showing the least tuning due to the vanishing
charge of the $H_u$ state. This model is quite interesting in the
sense that it provides a solution to the $\mu$-problem while avoiding
the large tuning (with current limits) from the $Z^\prime$
mass. However one should remember we are looking at conservative limits
on naturalness here and there is no solution to the usual tuning from
the large stops needed to get a $125$ GeV Higgs in this model.  In any
case future LHC limits on the $Z^\prime$ mass would be especially
interesting for this case as they would significantly raise the
minimum level of tuning which can be obtained.

>>>>

However in this study we will examine this source of fine tuning in more
detail by considering low energy constructions where the usual fine
tuning problem from the Higgs is minimised.  We will also consider
alternate charges for the extra $U(1)$ symmetry to relax the focus on
the E$_6$SSM and demonstrate that this is quite a generic result.

@@

= \underset{i}{\max}\,\{\Delta_{p_i}\}$.\\

@@

\begin{figure}[h]
\begin{center}
\resizebox{!}{6cm}{%
\includegraphics{mssm_stop_tuning.png}}
\resizebox{!}{6cm}{%
\includegraphics{mssm_higgs_tuning.png}}
\caption{Left panel: Scatter plot of fine tuning in the MSSM as a function of
the lightest stop mass, $m_{\tilde{t}_1}$, for the cut-off scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV. 
Right panel: Scatter plot of fine tuning in the MSSM as a function of
the lightest Higgs mass, $m_{h_1}$, for the cut-off scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV.}
\label{Fig:mssmstoptuning}
\end{center}
\end{figure}
To illustrate this, in the left panel of Fig.~\ref{Fig:mssmstoptuning}
we show the variation in fine tuning when we scan over the stop masses
and mixing, with $500 \textrm{ GeV} \leq m_{Q_3},m_{u_3}\leq
10\textrm{ TeV}$ and $-3810\textrm{ GeV}\leq A_t\leq -20\textrm{
  GeV}$, for $M_X=20$ TeV, $50$ TeV and $100$ TeV.  The remaining
parameters we fix such that at $M_{SUSY}$ they have the values
$\mu=-97.5$ GeV, $B=-84.8$ GeV, $M_1=92.1$ GeV, $M_2=95.9$ GeV,
$M_3=352$ GeV, $A_b=-117.9$ GeV, $A_\tau=-7.8$ GeV, $m_{L_i}=400$ GeV,
$m_{e_i}=204$ GeV, $m_{Q_{1,2}}=438$ GeV, $m_{u_{1,2}}436$ GeV and
$m_{d_i}=438$ GeV ($i=1,2,3$). Although we should stress that making
this choice will lead to a spectrum which is in conflict with the LHC
limits, doing so ensures that fine tuning due to the other parameters
is small, so that we avoid washing out the fine tuning impact of the
stops when the tuning is small\footnote{For models in which the
  spectrum is heavier, when the stop masses are small the fine tuning
  reaches a lower bound imposed by other heavier parameters.} as can
be the case when the stop masses are less than 1 TeV.  Note that the
Higgs mass is also allowed to vary in this scan, as shown in the right
panel of Fig.~\ref{Fig:mssmstoptuning}. This illustrates the tuning
problem which people have been worrying about since the discovery of
the 125 GeV Higgs boson as we see that raising the stop masses is also
pushing up the Higgs mass, meaning that heavier Higgs masses require
more fine tuning.  However for a low value of the UV scale this tuning
is not so severe unless the stops are very heavy, and a 125 GeV Higgs
can be obtained without much tuning in this unrealistic case where we
have minimised other sources of tuning.  On the other hand the tuning
becomes more severe as we increase the cutoff such that for $M_X =
10^{16}$ GeV a lightest stop masses of $1-3$ TeV can result in a fine
tuning of $\approx 100 - 1000$ and the minimum tuning we see consistent
with a $125$ GeV Higgs is $\approx 200$, as shown in
Fig.~\ref{Fig:mssmstoptuning}.

@@

Since the stop mass does not have such a large impact on the fine
tuning when the cutoff scale is very low we can use this to see more
clearly the impact of the $Z^\prime$ mass on fine tuning. To do so we
select a fixed low cutoff of $M_X=20$ TeV and compare the fine tuning
for two different values of the $Z^\prime$ mass. We choose to look at
$M_{Z^\prime}=2.5$ TeV, which is just above the current limits, and
$M_{Z^\prime}=4.5$ TeV, which should be in reach in run II at the LHC
\cite{}[CITE PLEASE] and then compare the fine tuning calculated in
each case to the tuning in the MSSM.  For this, we have performed a
large six dimensional parameter space scan in both the MSSM and
E$_6$SSM, varying those parameters most relevant for the fine tuning
and the Higgs mass.  Therefore the set of parameters which we vary
includes $\mu$, $B$ and $\tan \beta$ for the MSSM, and $\lambda$,
$A_\lambda$ and $\tan \beta$, for the E$_6$SSM, which appear at tree
level in the EWSB conditions of the models. While the RGE contribution
from large stop masses to the fine tuning is small for such a low
cut-off scale, the stop contributions to the effective potential can
play a significant role in reducing the fine tuning [WE SHOULD DISCUSS
  THIS].  For this reason it is still important to properly treat the
tuning associated with stop contributions to the one-loop effective
potential, and so we also scan over the soft masses $m_{Q_3}^2$,
$m_{u_3}^2$ and the stop mixing $A_t$.  The relevant parameters and
ranges that were scanned over are summarised in Table
\ref{tab:scanranges}. With the exception of the stop mixing $A_t$ and
the soft trilinear $A_\lambda$, which are logarithmically scanned, all
parameters are linearly scanned.  In addition to this we also repeat
each scan for three different values of $M_2$ to allow more variation
in the chargino masses.

@@

Nonetheless the MSSM the impact of potential chargino mass limits is
shown in Fig.~\ref{Fig:chargino-plot}.  There we see that if the the
full parameter space with $m_{\chi^\pm} < 700$ GeV was excluded the
impact would be to make the tuning in the MSSM with a $20$ TeV cutoff
slightly worse than that of the E$_6$SSM with the same cutoff and a
$Z^\prime$ mass just larger than current limits. In the E$_6$SSM,
while raising the chargino limit can have the same impact in
principle, due to current limits on the $Z^\prime$ mass already
imposing a significant degree of tuning, chargino masses do not make
much of a noticable change.

@@

Interestingly the coefficient $d$ is very small (and negative) in the
case of the $U(1)_I$.  This allows a dramatic reduction in the fine
tuning from the $U(1)_I$ symmetry.  This is a result of the $H_u$
charge associated with $U(1)_I$ vansihing, which means that the
D-terms to the lightest Higgs which is predominantly $H_u$ at large
$\tan \beta$ are suppressed, making it difficult to raise the Higgs
mass in the same way as happens in the other models and explaining why
heavier Higgs values in this model can't be obtained.  Therefore the fine
tuning behaviour in this model is closer to that of the MSSM, though
the LHC can still constrain the most natural scenarions from just
placing tougher limits on the $Z^\prime$ mass.

@@

To further illustrate this point we looked at how the tuning varies
with $M_X$ for low tuning benchmarks in the MSSM and E$_6$SSM. These
benchmarks are defined in table \ref{} and the results are shown in
Fig.~\ref{Fig:BMs-varyMX}. Since the behaviour is quite complicated we
now discuss these in detail as it provides some insight into the many
differences in the tuning between the two models.

@@

By contrast the tuning for E$_6$SSM BM2 is very different.  This point
was chosen as it had much lighter stops and a gluino mass that is just
above the experimental limit of $1.4$ TeV \cite{}.  At $20$ TeV it was
not amongst the lowest tuned points, since at that scale the tree
level tuning from $M_{Z^\prime}$ dominates.  However the reduction in
$M_3$ means that $\Delta_{M_3}$ is substantially lower and only
becomes the dominant tuning very close to the GUT scale, giving a
tuning at $10^{16}$ GeV of $\approx 900$, which is below that of the
other two benchmark points.  Since large mixing was also used to
reduce allow lighter stops, we see initially a large contribution from
$\Delta_{A_t}$ as we did in the MSSM benchmark.  [Last thing that
  needs to be explained - why do $\Delta_{M_3}$ and
  $\Delta_{m_{H_u}^2}$ start to decrease at large $M_X$.  I'm guessing
  something funny to do with their values and also $m_{Q}^2$ and $M_3$
  at the highscales.  but i need to see the numbers to know.]

@@

Prior to stringent experimental constraints on the mass of the
lightest Higgs boson and squarks in supersymmetric models, a simple
picture of a natural SUSY model emerged from theoretical reasoning,
with soft masses set to similar values at the GUT scale through local
gravity interactions with the hidden sector.  Through the use of
renormalisation group running one can then see that at the EW scale
the stops and soft Higgs masses (which appear directly in the EWSB
conditions) should appear in the the EWSB condition for $M_Z$ and
therefore it was expected that these masses should not be much bigger
than $100$ GeV.  However to disturb this elegant picture first LEP
placed constraints on the Higgs mass, requiring it to be above 114.4
GeV \cite{}, which already introduced significant tuning for
constrained models, since heavy stops are required to raised the
lightest Higgs mass above it's tree level bound. Recently this problem
got much worse since the LHC measured the Higgs to be around $125$
GeV.

@@

In contrast we find that in $U(1)$ extensions of the MSSM there is an
additional way to constrain the naturalness of the models, which is
through the $Z^\prime$ mass limit.  We find when we impose a low
cutoff of $20$ TeV for setting the soft masses, the lowest tuning in
the E$_6$SSM is compatible with a $Z^\prime$ mass of $2.5$ TeV was
$\Delta \approx 100$, while if the LHC run II can place a limit of
$4.5$ TeV on $M_Z^\prime$ then the tuning would be just under $400$.
By comparison the current situation in the MSSM only requires a tuning
of around $20-30$. [Dylan: update numbers a bit -- I suck at reading
  them off the graphs:)] This should be interpreted as saying that in
the most conservative limits one can place on naturalness on these
models, the tuning in the E$_6$SSM is worse.  However if there are no
charginos below $700$ GeV then the situation in the two models would
be the same.

This should also be contrasted with what happens as we raise the high
scale boundary, $M_X$.  We showed that for benchmark points that which
one is more tuned depends very strongly on $M_X$.  The E$_6$SSM tuning
is sufficiently complicated by the interplay of tehse different
sources of tesnion in the EWSB conditions that a small reduction in
fine tuning can even occur for a moderate increase in $M_X$.  However
as $M_X$ increases towards the scale where the gauge couplings unify,
the familiar tunings do dominate, though with tunings from the Gluino
mass appearing to be more significant relative to those from soft
scalar masses.

We also looked at the tuning in different $U(1)$ extensions for
fixed $\tan \beta = 10$.  We found that in all cases the fine tuning
was much worse for the larger $Z^\prime$ mass, further emphasising
the importance they play in these $U(1)$ extensions.  Interestingly
the actual level in tuning varied quite a bit between these models
with the $U(1)_I$ model showing the least tuning due to the vanishing
charge of the $H_u$ state. This model is quite interesting in the
sense that it provides a solution to the $\mu$-problem while avoiding
the large tuning (with current limits) from the $Z^\prime$
mass. However one should remember we are looking at conservative limits
on naturalness here and there is no solution to the usual tuning from
the large stops needed to get a $125$ GeV Higgs in this model.  In any
case future LHC limits on the $Z^\prime$ mass would be especially
interesting for this case as they would significantly raise the
minimum level of tuning which can be obtained.
